Mortality Trends Among Shelter Residents in Toronto from 2007 to 2024

### 2024-09-27 starter_folder-main/paper/paper.qmd
Grammar problems fix

---
title: "Mortality Trends Among Shelter Residents in Toronto from 2007 to 2024"
author: 
  - Fangning Zhang
thanks: "Code and data are available at: https://github.com/FangningZhang81/Deaths-of-Shelter-Residents."
date: today
date-format: long
abstract: "This study conducts a series of analyses of mortality among shelter residents in Toronto from 2007 to 2023, focusing on overall trends, gender differences, and seasonal variations. The results indicate a concerning increase in deaths starting from 2019, particularly among males. The number of deaths during the winter months is higher compared to other months. Understanding the mortality of shelter residents is crucial for formulating effective social policies, improving shelter services, and enhancing the health and well-being of vulnerable groups."
format: pdf
number-sections: true
bibliography: references.bib
toc: true
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
#hide the code in the PDF

@@

Although many existing researches highlight the higher mortality rates among the homeless population, there is still a lack of analysis specifically targeting shelter residents. In this study, a comprehensive analysis of mortality data is made among shelter residents in Toronto, focusing on the following key aspects. First, we analyze the overall trend in the number of deaths over the 16-year period from 2007 to 2023. Second, we explored gender differences in mortality. Third, we analyzed the monthly distribution of deaths to understand whether different climates or seasons have varying impacts on shelter residents.

The analysis revealed a concerning upward trend in the number of deaths starting from 2019, despite a slight decline in the following two years. Male deaths number have consistently been higher than female deaths, which is more and more significant after 2019. Additionally, there was an increase in deaths during the winter months, suggesting a link between extreme weather conditions and mortality risk may exist. This may consistent with previous research indicating that bad weather conditions and overcrowding in shelters can exacerbate health risks for homeless individuals (@vuillermoz2016mortality).

The rise in the number of deaths starting in 2019 worths attention. However, this increase does not necessarily indicate worsening shelter conditions or declining health among residents. The rise could be due to an increase in the number of shelter residents, changes in admission policies, or external factors such as the impact of COVID-19 and other social and public health issues (@ramin2009health). This is discussed in detail in the discussion section.

@@

The data was extracted, cleaned and visualize using R programming language and the whole coding process is done with R (@citeR). Some packages are used throughout the paper, including tidyverse (@cite_tidyverse), ggplot2 (@cite_ggplot), styler (@cite_styler), janitor (@citejanitor), OpenDataToronto
(@opendatatorotno), knitr (@cite_knitr), and KableExtra (@cite_kableExtra). Besides, the stater folder from Rohan (@rohan)is used.

@@

The raw data was first imported from the file and the column names were standardized using the `janitor` package(@citejanitor) to ensure consistency and ease of use throughout the analysis. The next step involved converting the month abbreviations into numeric values to increase the ease of use and analysis. For example, "01" for January, "02" for February.

A new column named `year_month` was generated by concatenating the `year` and `month` columns making it easier to perform time-series analyses and visualizations.

The dataset also included a column named `transgender_non_binary_two_spirit`, which some rows contained non-numeric values (n/a). All `NA` values in this column were replaced with 0 to ensure the data was suitable for statistical analysis and visualization.

Finally, the cleaned dataset was saved to a new file named `analysis_data.csv` in the `analysis_data` directory. It includes the following variables:

@@

The @tbl-summary below provides a summary of the cleaned data, showing key statistics (maximum, minimum, mean, median, and variance) for the total number of deaths, as well as for male and female decedents. It highlights that the average number of deaths per month is 4, with males averaging 3 deaths and females 1 death.

@@

```{r}
#| warning: false
#| message: false
#| label: fig-total
#| fig-cap: "Total Number of Deaths by Year (2007-2023)"
# Filter data for the years 2007 to 2023
filtered_data <- data |>
  filter(year >= 2007 & year <= 2023)

@@

```
@fig-category displays the death trend of each category among shelter residents in Toronto from 2007 to 2023. The data shows a significant increase in deaths starting from 2019 of all three lines, while the yellow line remains at a relatively low level. The total number of deaths equals the sum of male, female, and other gender deaths.The males death rate has consistently been higher than females and the differences is becoming increasingly obvious. The pie chart (@fig-dis) shows the gender distribution of the deaths. 78.2% of the total deaths are from males, while females account for 65.3%. In contrast, individuals of other genders make up only 4.7%.

@@

# Plot the bar chart
ggplot(monthly_deaths, aes(x = factor(month), y = total_deaths)) +
  geom_bar(stat = "identity", fill = "lightblue") +
  labs(
    x = "Month",
    y = "Total Number of Deaths"
  ) +
  scale_x_discrete(labels = 1:12) +
  geom_text(aes(label = total_deaths), # Set x-axis from 1 to 12
            vjust = -0.5,    
            size = 3,        
            color = "black") +
  theme_minimal() 
```
@fig-month illustrates the total number of deaths by month among shelter residents. The highest number of deaths occurred in January which is 80 people, followed by relatively high numbers in November which is 72 people. Conversely, the lowest number of deaths occurred in March and September which is 53 and 51 separatly. There is a slight increase in deaths during the winter months (December to February).

## Other Datasets
There are some related datasets found on OpenDataToronto (@opendatatorotno), but they are not used for the following reasons.The "Death Registry Statistics" dataset contains variables that are too broad and do not provide specific information on deaths among vulnerable populations, making it unsuitable for detailed analysis of shelter resident mortality.The "Fatal and non-fatal suspected opioid overdoses in the shelter system" dataset, while relevant, only covers one specific cause of death (opioid overdoses) and therefore does not capture the full scope of mortality within the shelter system, making it insufficient as the primary dataset. The dataset "Daily Shelter Occupancy" can be a very helpful dataset for this study. As The number of deaths among shelter residents is closely related to the total number of people residing in the shelter. The "Daily Shelter Occupancy" dataset could be useful for understanding the relationship between death rates and the total number of shelter residents, but it was not included in this study because the primary focus is on the death statistics themselves, rather than occupancy rates. Future studies could consider integrating this dataset to explore such relationships in more depth.

@@

The study revealed an overall upward trend in the mortality among shelter residents, with a significant gender difference which is the death number for males is consistently higher than that for females. Additionally, there is a more deaths during the winter months. A particularly concerning finding is the sharp rise in mortality starting in 2019.

>>>>

---
title: "Mortality Trends Among Shelter Residents in Toronto from 2007 to 2024"
author: 
  - Fangning Zhang
thanks: "Code and data are available at: https://github.com/FangningZhang81/Deaths-of-Shelter-Residents."
date: today
date-format: long
abstract: "This study conducts a series of analyses of mortality among shelter residents in Toronto from 2007 to 2023, focusing on overall trends, gender differences, and seasonal variations. The results indicate a concerning increase in deaths starting from 2019, particularly among males. The number of deaths during the winter months is higher compared to other months. Understanding the mortality of shelter residents is crucial for formulating effective social policies, improving shelter services, and enhancing the health and well-being of this vulnerable group."
format: pdf
number-sections: true
bibliography: references.bib
toc: true
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
#hide the code in the PDF

@@

Although many existing researches highlight the higher mortality rates among the homeless population, there is still a lack of analysis specifically targeting shelter residents. In this study, a comprehensive analysis of mortality data is made among shelter residents in Toronto, focusing on the following key aspects. First, we analyze the overall trend in the number of deaths over the 16-year period from 2007 to 2023. Second, we explore gender differences in mortality. Third, we analyze the monthly distribution of deaths to understand whether different climates or seasons have varying impacts on shelter residents.

The analysis reveal a concerning upward trend in the number of deaths starting from 2019, despite a slight decline in the following two years. Male deaths number have consistently been higher than female deaths, which is more and more significant after 2019. Additionally, there is an increase in deaths during the winter months, suggesting a link between extreme weather conditions and mortality risk may exist. This may consistent with previous research indicating that bad weather conditions and overcrowding in shelters can exacerbate health risks for homeless individuals (@vuillermoz2016mortality).

The rise in the number of deaths starting in 2019 deserves attention. However, this increase does not necessarily indicate worsening shelter conditions or declining health among residents. The rise could be due to an increase in the number of shelter residents, changes in admission policies, or external factors such as the impact of COVID-19 and other social and public health issues (@ramin2009health). This is discussed in detail in the discussion section.

@@

The data is extracted, cleaned and visualized using R programming language and the whole coding process is done with R (@citeR). Some packages are used throughout the paper, including tidyverse (@cite_tidyverse), ggplot2 (@cite_ggplot), styler (@cite_styler), janitor (@citejanitor), OpenDataToronto
(@opendatatorotno), knitr (@cite_knitr), and KableExtra (@cite_kableExtra). Besides, the stater folder from Rohan (@rohan)is used.

@@

The raw data is first imported from the file and the column names are standardized using the `janitor` package(@citejanitor) to ensure consistency and ease of use throughout the analysis. The next step is converting the month abbreviations into numeric values to increase the ease of use and analysis. For example, "01" for January, "02" for February.

A new column named `year_month` is generated by concatenating the `year` and `month` columns making it easier to perform time-series analyses and visualizations.

The dataset also includes a column named `transgender_non_binary_two_spirit`, which contain non-numeric values (n/a) in some rows. All `NA` values in this column are replaced with 0 to ensure the data is suitable for statistical analysis and visualization.

Finally, the cleaned dataset is saved to a new file named `analysis_data.csv` in the `analysis_data` directory. It includes the following variables:

@@

The @tbl-summary below provides a summary of the cleaned data, showing key statistics (maximum, minimum, mean, median, and variance) for the total number of deaths, as well as for male and female decedents. It highlights the average number of deaths per month is 4, with males averaging 3 deaths and females 1 death.

@@

```{r}
#| warning: false
#| message: false
#| label: fig-total
#| fig-cap: "Total Number of Deaths by Year (2007-2023)"

# Filter data for the years 2007 to 2023
filtered_data <- data |>
  filter(year >= 2007 & year <= 2023)

@@

```
@fig-category displays the death trend of each category among shelter residents in Toronto from 2007 to 2023. The data shows a significant increase in deaths starting from 2019 of all three lines, while the yellow line remains at a relatively low level. The total number of deaths equals the sum of male, female, and other gender deaths.The males death rate has consistently been higher than females and the differences are becoming increasingly obvious. The pie chart (@fig-dis) shows the gender distribution of the deaths. 78.2% of the total deaths are from males, while females account for 65.3%. In contrast, individuals of other genders make up only 4.7%.

@@

# Plot the bar chart
ggplot(monthly_deaths, aes(x = factor(month), y = total_deaths)) +
  geom_bar(stat = "identity", fill = "lightblue") +
  labs(
    x = "Month",
    y = "Total Number of Deaths"
  ) +
  scale_x_discrete(labels = 1:12) +
  geom_text(aes(label = total_deaths), # Set x-axis from 1 to 12
            vjust = -0.5,    
            size = 3,        
            color = "black") +
  theme_minimal() 
```
@fig-month illustrates the total number of deaths by month among shelter residents. The highest number of deaths occurred in January which is 80 people, followed by relatively high numbers in November which is 72 people. Conversely, the lowest number of deaths occurred in March and September which is 53 and 51 separately. There is a slight increase in deaths during the winter months (December to February).

## Other Datasets
There are some related datasets found on OpenDataToronto (@opendatatorotno), but they are not used for the following reasons.The "Death Registry Statistics" dataset contains variables that are too broad and do not provide specific information on deaths among vulnerable populations, making it unsuitable for detailed analysis of shelter resident mortality.The "Fatal and non-fatal suspected opioid overdoses in the shelter system" dataset, while relevant, only covers one specific cause of death (opioid overdoses) and therefore does not capture the full scope of mortality within the shelter system, making it insufficient as the primary dataset. The dataset "Daily Shelter Occupancy" can be a very helpful dataset for this study. As The number of deaths among shelter residents is closely related to the total number of people residing in the shelter. This dataset could be useful for understanding the relationship between death rates and the total number of shelter residents, but it was also not included in this study because the primary focus is on the death statistics themselves, rather than occupancy rates. Future studies could consider include this dataset to explore such relationships.

@@

The study reveal an overall upward trend in the mortality among shelter residents, with a significant gender difference which is the death number for males is consistently higher than that for females. Additionally, there is a more deaths during the winter months. A particularly concerning finding is the sharp rise in mortality starting in 2019.
